Measurement Clusterization in Bayesian D-optimal Designs in Infinite Dimensions

### 2024-08-11 latex/intro.tex
done

\item \label{q:why} \textbf{Why does measurement clusterization
  occur?} In Section \ref{subsec:why} (specifically, Fig~\ref{fig:why}
  and the discussion following it) we give a compelling answer to this
  question by transporting insights we gain from our generic model to
  the inverse problem of the 1D heat equation. Our analysis reveals
  that a D-optimal design focuses on a select set of prior
  eigenvectors, specifically those with the largest eigenvalues in the
  prior covariance spectrum. In practical scenarios, the number of
  locations where (a) all relevant prior eigenvectors are
  significantly large, and (b) other eigenvectors are close to zero,
  is limited. Consequently, the clusterization of measurements arises
  as a natural consequence of the pigeonhole principle, as there are
  more measurements available than there are locations satisfying
  conditions (a) and (b).

>>>>

\item \label{q:why} \textbf{Why does measurement clusterization
  occur?} In Sections \ref{subsec:why} we give two compelling answers
  to this question by (i) transporting insights we gain from our
  generic model to the inverse problem of the 1D heat equation, and
  (ii) connecting measurement clusterization to Carath\'eodory's
  Theorem.

  Our analysis for the heat equation relies on conclusions from our
  generic model. In particular, Theorem \ref{thm:char} reveals that a
  D-optimal design focuses on a select set of prior eigenvectors,
  i.e.~those with the largest eigenvalues in the prior covariance
  spectrum. In practical scenarios, the number of locations where (a)
  all relevant prior eigenvectors are significantly large, and (b)
  other eigenvectors are close to zero, is limited. Consequently, the
  clusterization of measurements arises as a natural consequence of
  the pigeonhole principle, as there are more measurements available
  than there are locations satisfying conditions (a) and (b).

  The connection to Carath\'eodory's Theorem also utilizes the
  concentration of measurement effort on the abovementioned select set
  of prior eigenvectors. This concentration allows us to move from an
  infinite-dimensional setting to finite dimensions, where the
  conditions of Carath\'eodory's Theorem hold. We conclude that a
  D-optimal design arises by weighting a small number of
  measurements. If the number of allowed measurements is larger than
  the small number dictated by Carath\'eodory's Theorem, we have
  excessive weight on some measurements, which can be interpreted as
  clusterization.

### 2024-08-11 latex/vanishing.tex
done

\subsubsection{Clusterization in our generic model}
We can gain further insight to clusterization in our generic model,
from Carath\'eodory's Theorem and the concentration on the first $k$
eigenvectors of $\fwd \prcov \fwd^*$. We present a short discussion
adapting arguments presented in \cite[Chapter
  3]{silveyOptimalDesign1980} and \cite[Section
  5.2.3]{pronzatoPazman2013}.

Consider $\opt$ a D-optimal design under our generic model.  As
instructed by Theorem \ref{thm:char}, we ignore all but the first $k$
eigenvectors of $\fwd \prcov \fwd$. Thus, we replace $\hilo$ with
$\hilo^{(k)}$ --- the $k$-dimensional subspace spanned by the first
$k$ eigenvectors of $\fwd^*\prcov\fwd$. We keep in mind that
$\hilo^{(k)}$ is isomorphic to $\mathbb{R}^k$ and let
\begin{equation*}
  \mathcal{M} := \conv \{\meas \meas^* : \meas\in \hilo^{(k)}, \|\meas\|=1\},
  %\hilo^{(k)}\},
\end{equation*}
where $\conv$ denotes the convex hull of a set. The set $\mathcal{M}$
contains only positive-definite matrices, hence it lives in a
$k(k+1)/2$-dimensional vector space. Since $\opt^*\opt = \sum_{i=1}^m
\meas_i\meas_i^*$ for $\meas_i \in \hilo^{(k)}$, it is easy to verify
that $\frac1m \opt^*\opt \in \mathcal{M}$.  Now, recall
Carath\'eodory's Theorem:
\begin{theorem}[Carath\'eodory]
  Let $X \subseteq \mathbb{R}^n, X \neq \phi$ and denote $\conv (X)$
  the convex hull of $X$. For every $x \in \conv (X)$, $x$ is a convex
  combination of at most $n+1$ vectors in $X$.
\end{theorem}
Carath\'eodory's Theorem implies that there exist $\meas_i$ and
$\alpha_i$ such that
\begin{equation*}
  \opt^*\opt = \sum_{i=1}^I \alpha_i \meas_i\meas_i^*,
\end{equation*}
where $\|\meas\|=1, \sum\alpha_i = m, \alpha_i \geq 0$ and $I =
\frac{k(k+1)}{2} + 1$. We can thus write $\opt$ as (note that we do
not require $\meas_i$ to be orthogonal to each other):

\[
\opt =
\left[
  \begin{array}{ccc}
    \horzbar & \sqrt{\alpha_i} \meas^*_1 & \horzbar \\
    \horzbar & \sqrt{\alpha_2} \meas_2^* & \horzbar \\
             & \vdots    &          \\
    \horzbar & \sqrt{\alpha_I} \meas_I^* & \horzbar \\
  \end{array}
\right]
\]

Therefore, if $m > \frac{k(k+1)}{2} + 1$, then at least one
coefficient $\alpha_i > 1$. Note that the abovementioned $\opt$ is not
a valid design (since its rows do not have unit norm,
i.e.~$\|\sqrt{\alpha_i}\meas\| \neq 1$). However, such representation
provides another, deep view of clusterization.

%% For example, if $k=2$, there exists an optimal design $\opt$ that can
%% be represented be as a linear combination of at most $(2+1)/2 + 1 = 4$
%% measurement vectors $\meas$, with nonnegative coefficients whose
%% squares sum to $m$. If, the number of allowed measurements $m=5$, we
%% can write $\opt = \sum_{i=1}^4 \alpha_i \meas_i$, where $\alpha_i > 1$
%% for some $i\in \{1,2,3,4\}$. Since we put weight $>1$ on the
%% $i\textsuperscript{th}$ measurement, we can view such a design as
%% clustered.

%%  whereThus, if we view $\obs$ as a $\mathbb{R}^{m\times k}$ matrix, we
%%  conclude that $\obs^*\obs$ is a symmetric positive definite matrix
%%  that lives in a $k(k+1)/2$-dimensional space.

%% The set $\conv \mathcal{M}$ is a superset of all possible

%% Since we restrict $\|\meas\| = 1$, the set of allowed
%% $\obs^*\obs$ is the ball of norm $m$ unit ball in (a space isomorphic to)
%% $\mathbb{R}^k.

%% $\{\meas \meas: \meas \in \span (\ev_1,\ev_2,\dots,
%% \ev_k\}$ is in fact $k$-dimensional.

%%   By Theorem
%%   \ref{thm:char}, we ignore all eigenvectors except for the first
%%   $k$.\ \

%% First, note that $\fwd^*\obs^*\obs\fwd$ is the Fisher inf

>>>>

\subsubsection{Clusterization in our generic model}%\label{subsub:cara}
We can gain further insight to clusterization in our generic model,
from Carath\'eodory's Theorem and the concentration on the first $k$
eigenvectors of $\fwd \prcov \fwd^*$. We present a short discussion
adapting arguments presented in \cite[Chapter
  3]{silveyOptimalDesign1980} and \cite[Section
  5.2.3]{pronzatoPazman2013}.

Consider $\opt$ a D-optimal design under our generic model.  As
instructed by Theorem \ref{thm:char}, we ignore all but the first $k$
eigenvectors of $\fwd \prcov \fwd$. Thus, we replace $\hilo$ with
$\hilo^{(k)}$ --- the $k$-dimensional subspace spanned by the first
$k$ eigenvectors of $\fwd^*\prcov\fwd$. Let
\begin{equation*}
  \mathcal{M} := \conv \{\meas \meas^* : \meas\in \hilo^{(k)}, \|\meas\|=1\},
  %\hilo^{(k)}\},
\end{equation*}
where $\conv$ denotes the convex hull of a set. The set $\mathcal{M}$
contains only positive-definite operators on a $k$-dimensional vector
space. Hence $\mathcal{M}$ lives in a $k(k+1)/2$-dimensional vector
space. Since $\opt^*\opt = \sum_{i=1}^m \meas_i\meas_i^*$ for $\meas_i
\in \hilo^{(k)}$, it is easy to verify that $\frac1m \opt^*\opt \in
\mathcal{M}$.  Recall Carath\'eodory's Theorem:
\begin{theorem}[Carath\'eodory]
  Let $X \subseteq \mathbb{R}^n, X \neq \phi$ and denote $\conv (X)$
  the convex hull of $X$. For every $x \in \conv (X)$, $x$ is a convex
  combination of at most $n+1$ vectors in $X$.
\end{theorem}
Carath\'eodory's Theorem implies that there exist $\meas_i$ and
$\alpha_i$ such that
\begin{equation*}
  \opt^*\opt = \sum_{i=1}^I \alpha_i \meas_i\meas_i^*,
\end{equation*}
where $\|\meas\|=1, \sum\alpha_i = m, \alpha_i \geq 0$ and $I =
\frac{k(k+1)}{2} + 1$. We can thus write $\opt$ as:
%(note that we do not require $\meas_i$ to be orthogonal to each
%other):

\[
\opt =
\left[
  \begin{array}{ccc}
    \horzbar & \sqrt{\alpha_i} \meas^*_1 & \horzbar \\
    \horzbar & \sqrt{\alpha_2} \meas_2^* & \horzbar \\
             & \vdots    &          \\
    \horzbar & \sqrt{\alpha_I} \meas_I^* & \horzbar \\
  \end{array}
\right].
\]

Unfortunately, $\opt$ is not a valid design, since its rows do not
have unit norm. Still, the above representation of $\opt$ is useful:
If $m > \frac{k(k+1)}{2} + 1$, then $\alpha_i > 1$ for some $1\leq i
\leq I$.  Thus, we can view $\opt$ as a clustered design, since it
places weight $>1$ on a single measurement vector.
